---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
author: "Upasana Sharma"
date: "3/24/2022"
output: pdf_document
editor_options: 
chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)

```

```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth))
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n))
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})

```

# Solution for Task 1

```{r 0 Load Libraries, results = 'hide'}
#### Loading all necessary libraries
library(data.table)
library(readr)
library(ggplot2)
library(ggmosaic)

#### Opening dataset files
filePath <- "C:/Quantium/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
```

## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look at each of the datasets provided.
### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the data. As we have read in the dataset as a `data.table` object, we can also run
`transactionData` in the console to see a sample of the data or use `head(transactionData)` to look at the first 10 rows.

Let's check if columns we would expect to be numeric are in numeric form and date columns are in date format.

```{r Examining transaction data}
#### View the structure of transactionData table
str(transactionData)
head(transactionData)
```

We can see that the date column is in an integer format. Let’s change this to a date format.

```{r}
#### Converting the date from integer format to date format
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
head(transactionData)
```

Lets check that we are looking at the right products by taking a closer look at PROD_NAME.

```{r}
#### Checking the column PROD_NAME by its summary
transactionData[, .N, by = PROD_NAME]
```

Looks like we are definitely looking at potato chips but how can we check that these are all chips? We can do some basic text analysis by summarizing the individual words in the product name.

```{r}
#### Checking for incorrect entries other than chips
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), "
")))
setnames(productWords, 'words')
```

As we are only interested in words that will tell us if the product is chips or not, let's remove all words with digits and special characters such as '&' from our
set of product words.

```{r}
#### Filtering data for digits and special characters
# Filtering data for digits and special characters

# Removing digits
productWords <- productWords[grepl("\\d", words) == FALSE, ]

# Removing special characters
productWords <- productWords[grepl("[:alpha:]", words), ]

# Let's look at the most common words by counting the number of times a word appears and sorting them by this frequency in order of highest to lowest frequency
productWords[, .N, words][order(N, decreasing = TRUE)]

# Check the data
head(transactionData)
```

There are salsa products in the dataset but we are only interested in the chips category, so let's remove these.

```{r}
#### Remove salsa from the items list
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max values for each feature to see if there are any obvious outliers in the data and if there are any nulls in any of the columns (`NA's : number of nulls` will appear in the output if there are any nulls).

```{r}
summary(transactionData)

#### Check for nulls
transactionData[is.null(PROD_NAME), .N]
```

There are no nulls in the columns but product quantity appears to have an outlier which we should investigate further. Let's investigate further the case where 200
packets of chips are bought in one transaction.

```{r}
#### Filter the data to find the outlier 
transactionData[PROD_QTY == 200]
```

There are two transactions where 200 packets of chips are bought in one transaction and both of these transactions were by the same customer.

```{r}
#### See if this customer has had any other transactions
transactionData[LYLTY_CARD_NBR == 226000]
```

It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#### Filtering the customer based on the loyalty card number
transactionData = filter(transactionData, LYLTY_CARD_NBR != 226000)

#### Re-examine the data
summary(transactionData)
```

That's better. Now, let's look at the number of transaction lines over time to see if there are any obvious data issues such as missing data.

```{r}
# Count the number of transaction by date
transactionData[, .N, by = DATE]

# Creating a sequence of dates and join this the count of transactions by date
transactions_by_date <- as.data.table(transactionData)
transactions_by_date <- data.frame(transactions_by_date[, .N, by = DATE])
```

There's only 364 rows, meaning only 364 dates which indicates a missing date. 
Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.

```{r}
#### Setting plot theme to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#### Plot transactions over time
ggplot(transactions_by_date, aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

We can see that there is an increase in purchases in December and a break in late
December. Let's zoom in on this.

```{r}
#### Filter to December and look at individual days
ggplot(transactions_by_date[transactions_by_date$DATE >= "2018-12-10" & transactions_by_date$DATE <= "2018-12-31", ], aes(x = DATE, y = N)) +
  geom_line() +
  labs(x = "Date", y = "Number of transactions", title = "Transactions over time") +  scale_x_date(breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on
Christmas day.
Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.

```{r}
#### Create pack size
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]

#### Checking for any outliers
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]

# Let's check the output of the first few rows to see if we have indeed picked out pack size.
transactionData
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r}
#### Plotting a histogram showing the number of transactions by pack size
hist(transactionData[, PACK_SIZE])
```

Pack sizes created look reasonable.
Now to create brands, we can use the first word in PROD_NAME to work out the brand name...

```{r}
#### Creating a column of brand names from the first word of column Product names
transactionData$BRAND <- gsub("([A-Za-z]+).*", "\\1", transactionData$PROD_NAME)
View(transactionData)

#### Identify unique brand names
transactionData[, .N, by = BRAND][order(-N)]
```

Some of the brand names look like they are of the same brands - such as RED and RRD, which are both Red Rock Deli chips. Let's combine these together.

```{r}
#### Clean Brand Names
transactionData[BRAND == "RED", BRAND := "RRD"]
transactionData[BRAND == "SNBTS", BRAND := "SUNBITES"]
transactionData[BRAND == "INFZNS", BRAND := "INFUSIONS"]
transactionData[BRAND == "WW", BRAND := "WOOLWORTHS"]
transactionData[BRAND == "SMITH", BRAND := "SMITHS"]
transactionData[BRAND == "NCC", BRAND := "NATURAL"]
transactionData[BRAND == "DORITO", BRAND := "DORITOS"]
transactionData[BRAND == "GRAIN", BRAND := "GRNWVES"]

#### Checking Again
transactionData[, .N, by = BRAND][order(-N)]
```

### Examining customer data
Now that we are happy with the transaction dataset, let's have a look at the customer dataset.

```{r}
#### View the structure of customerData table
str(customerData)
summary(customerData)
head(customerData)

#### Checking customer data by lifestage column
customerData[, .N, by = LIFESTAGE][order(N)]

#### Plotting a graph 
ggplot(customerData[, .N, by = LIFESTAGE][order(N)], aes(x = LIFESTAGE, y = N)) +
  geom_bar(stat = "identity", fill = "blue") + labs(x= "Life Stage", y = "Number of Customers", title = "Number of customers on each lifestage") + 
  theme(axis.text.x = element_text(angle = 55, vjust = 0.5))

#### Checking customer data by permium customers column
customerData[, .N, by = PREMIUM_CUSTOMER][order(N)]

#### Plotting a graph
ggplot(customerData[, .N, by = PREMIUM_CUSTOMER][order(N)], aes(x = PREMIUM_CUSTOMER, y = N)) +
  geom_bar(stat = "identity", fill = "blue") + labs(x= "Budget type", y = "Number of Customers", title = "Number of customers on each budget type") + 
  theme(axis.text.x = element_text(angle = 55, vjust = 0.5))

#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
View(data)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be sure that no duplicates were created. This is because we created `data` by setting `all.x = TRUE` (in other words, a left join) which means take all the rows in `transactionData` and find rows with matching values in shared columns and then joining the details in these rows to the `x` or the first mentioned table.

```{r}
#### Checking for Nulls in data
data[is.null(LIFESTAGE), .N]
data[is.null(PREMIUM_CUSTOMER), .N]
```

Great, there are no nulls! So all our customers in the transaction data has been accounted for in the customer dataset.

Data exploration is now complete!

## Data analysis on customer segments
Now that the data is ready for analysis, we can define some metrics of interest to
the client:
- Who spends the most on chips (total sales), describing customers by lifestage and
how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment
We could also ask our data team for more information. Examples are:
- The customer's total spend over the period and total spend for each transaction
to understand what proportion of their grocery spend is on chips
- Proportion of customers in each customer segment overall to compare against the
mix of customers who purchase chips
Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and
plotting the split by these segments to describe which customer segment contribute
most to chip sales.

```{r}
#### calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER
sales <- data[, .(SALES = sum(TOT_SALES)), .(LIFESTAGE, PREMIUM_CUSTOMER)]

#### Plotting a graph
ggplot(data = sales) +
  geom_mosaic(aes(weight = SALES, x = product(PREMIUM_CUSTOMER, LIFESTAGE),
  fill = PREMIUM_CUSTOMER)) + 
  labs(x = " Lifestage", y = "Premium Customers", title = "Proportion of sales") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Sales are coming mainly from Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees
Let’s see if the higher sales are due to there being more customers who buy chips.

```{r}
#### Calculating number of customers by LIFESTAGE and PREMIUM_CUSTOMER
customers <- data[, .(CUSTOMERS = uniqueN(LYLTY_CARD_NBR)), .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-CUSTOMERS)]

#### Plotting a graph
ggplot(data = customers) +
  geom_mosaic(aes(weight = CUSTOMERS, x = product(PREMIUM_CUSTOMER, LIFESTAGE),
                  fill = PREMIUM_CUSTOMER)) + 
  labs(x = " Lifestage", y = "Premium Customers", title = "Proportion of customers") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy chips. This contributes to there being more sales to these customer segments but this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.

Let's have a look at this next.


```{r}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
avg_units <- data[, .(AVERAGE = sum(PROD_QTY) / uniqueN(LYLTY_CARD_NBR)), .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVERAGE)]

#### Plotting a graph
ggplot(data = avg_units, aes(weight = AVERAGE, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge()) +
  labs(x = "Lifestage", y = "Avg units per transaction", title = "Units per customer") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Older families and young families in general buy more chips per customer

Let's also investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.

```{r}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
avg_price <- data[, .(AVERAGE = sum(TOT_SALES) / sum(PROD_QTY)), .(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVERAGE)]

#### Plotting a graph
ggplot(data = avg_price, aes(weight = AVERAGE, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge()) +
  labs(x = "Lifestage", y = "Avg price per transaction", title = "Price per unit") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than 
their own consumption. This is also supported by there being fewer premium midage and young singles and couples buying chips compared to their mainstream
counterparts.

As the difference in average price per unit is not large, we can check if this difference is statistically significant. To do so, I will perform an independent
t-test between mainstream vs premium midage young singles and couplesto see if thedifference is significant.Our data will yield revelant results about the
statistical significance of the price diference. we have all the data which uses ordinal scale as measurement applied to the data, if we plot price data 
it result on a normal distribution, we can rely on the results with reasonable assurance.

```{r}
priceperunit <- data[, price := TOT_SALES / PROD_QTY]
```

Let's proceed with the t-test

```{r}
#### Performing an independent t-test between mainstream vs premium and budget midage and young singles and couples
t.test(data[LIFESTAGE %in% c("YOUNG SINGELS/COUPLES", "MIDAGE SINGLES/COUPLES")
            & PREMIUM_CUSTOMER == "Mainstream", price]
       , data[LIFESTAGE %in% c("YOUNG SINGELS/COUPLES", "MIDAGE SINGLES/COUPLES")
            & PREMIUM_CUSTOMER != "Mainstream", price], 
       alternative = "greater")
```

The t-test results in a p-value < 2.2e-16, i.e. the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples.

## Deep dive into specific customer segments for insights
We have found quite a few interesting insights that we can dive deeper into.

We might want to target customer segments that contribute the most to sales to retain them or further increase sales. Let's look at Mainstream - young singles/couples. For instance, let's find out if they tend to buy a particular brand of chips.

```{r}
#### Deep dive into Mainstream, young singles/couples
segment_1 <- data[LIFESTAGE == "YOUNG SINGELS/COUPLES" & PREMIUM_CUSTOMER == "Mainstream", ]
others <- data[!(LIFESTAGE == "YOUNG SINGELS/COUPLES" & PREMIUM_CUSTOMER == "Mainstream"), ]

#### Comparing brand affinity to the rest of the population
quantity_segment_1 <- segment_1[, sum(PROD_QTY)]
quantity_others <- others[, sum(PROD_QTY)]
quantity_segment_1_by_brand <- segment_1[, .(targetSegment = sum(PROD_QTY) / quantity_segment_1), by = BRAND]
quantity_others_by_brand <- others[, .(others = sum(PROD_QTY) / quantity_others), by = BRAND]
brand_proportions <- merge(quantity_segment_1_by_brand, quantity_others_by_brand)[, affinityToBrand := targetSegment / others]
brand_proportions[order(-affinityToBrand)]
```

We can see that :

• Mainstream young singles/couples are 23% more likely to purchase Tyrrells chips compared to the
rest of the population.

• Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest
of the population

Let’s also find out if our target segment tends to buy larger packs of chips.

```{r}
#### Preferred pack size compared to the rest of the population
quantity_segment_1_by_pack <- segment_1[, .(targetSegment = sum(PROD_QTY) / quantity_segment_1), by = PACK_SIZE]
quantity_others_by_pack <- others[, .(others = sum(PROD_QTY) / quantity_others), by = PACK_SIZE]
pack_proportions <- merge(quantity_segment_1_by_pack, quantity_others_by_pack)[, affinityToPack := targetSegment / others]
pack_proportions[order(-affinityToPack)]
```

It looks like Mainstream young singles/couples are 27% more likely to purchase a 270g pack of chips compared to the rest of the population but let’s dive into what brands sell this pack size.

```{r}
data[PACK_SIZE == 270, unique(PROD_NAME)]
```

Twisties are the only brand offering 270g packs and so this may instead be reflecting a higher likelihood of purchasing Twisties.

Conclusion:

Let’s recap what we’ve found!

Sales have mainly been due to Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees shoppers. We found that the high spend in chips for mainstream young singles/couples and retirees is due to there being more of them than other buyers. Mainstream, midage and young singles and couples are also more likely to pay more per packet of chips. This is indicative of impulse buying behaviour. We’ve also found that Mainstream young singles and couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population. The Category Manager may want to increase the category’s performance by off-locating some Tyrrells and smaller packs of chips in discretionary space near segments where young singles and couples frequent more often to increase visibilty and impulse behaviour.




